---
title: "Data visualization: categorical data"
author: "<br>CMSACamp 2026<br><br>Department of Statistics & Data Science<br>Carnegie Mellon University"
footer:  "[36-SURE.github.io/2026](https://36-sure.github.io/2026)"
format:
  revealjs:
    theme: theme.scss
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.5
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| eval: true
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
# ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```

# Background

## Data

- Our dataset is **Wimbledon women's tennis shot-level data**

- The original data is publicly available on the [SCORE Sports Data Repository](https://data.scorenetwork.org/tennis/tennis-shot-level-data.html)

   - We will look at a *random sample of 10,000 (non-serve) shots*
   
   - The data was pre-processed slightly for this lecture:

```{r}
#| eval: false
# set seed for reproducibility
set.seed(4747)

# import data from current working directory
tennis_shots <- read_csv("tennis-w-shots-wim.csv.gz") |>
  # tidy names
  janitor::clean_names() |>
  # remove serves
  filter(shot_type != "serve") |>
  # NA = shot did not produce an outcome (i.e., returned)
  mutate(outcome_type = if_else(is.na(outcome_type),"returned", outcome_type)) |>
  # replace "_" with " " in all character variables 
  mutate(across(where(is.character), ~ str_replace_all(., "_", " "))) |>
  # convert date into ymd format
  mutate(date = ymd(date)) |>
  # random sample of sample 10,000 observations
  dplyr::slice_sample(n = 10000)
```

## Data

Let's begin by importing the data:

```{r}
library(tidyverse)
theme_set(theme_light()) # setting the ggplot theme

tennis_shots <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/tennis_w_shots_wim.csv") |>
  # filter out NAs for shot_type and outcome_type
  filter(!is.na(shot_type), !is.na(outcome_type)) |>
  # convert characters into factors
  mutate(across(where(is.character), as.factor)) 
```

. . .

<br>

*What does each row represent here?*


```{r}
names(tennis_shots)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```

## Categorical data

Two different versions of categorical data:

**Nominal**: categorical variables having unordered scales

-   Examples: race, gender, species, etc.

**Ordinal**: ordered categories; levels with a meaningful order

-   Examples: education level, grades, ranks, etc.

## Factors in `R`

-   In `R`, factors are used to work with categorical variables

-   `R` treats factors as ordinal - defaults to alphabetical

    -   May need to manually define the factor levels (e.g., the reference level)

-   See the [`forcats`](https://forcats.tidyverse.org/) package (automatically loaded with `tidyverse`)


```{r}
class(tennis_shots$shot_type)
levels(tennis_shots$shot_type)
```

# 1D categorical data

## Summarizing 1D categorical data

*How often is each shot type used during women's singles at the Wimbledon Championships?*

**Frequency tables** (counts)


```{r}
table(tennis_shots$shot_type)
```


```{r}
#tennis_shots |> 
#  group_by(shot_type) |>
#  summarize(n = n(), .groups = "drop")

tennis_shots |> 
  count(shot_type)
```

## Summarizing 1D categorical data

**Proportion table**

```{r}
prop.table(table(tennis_shots$shot_type))

tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n))
```

## Visualizing 1D categorical data

::: columns
::: {.column width="50%" style="text-align: left;"}
Create a bar chart with `geom_bar()`

-   Map `shot_type` to the x-axis

-   Counts of each category are displayed on the y-axis

```{r}
#| eval: false
tennis_shots |> 
  ggplot(aes(x = shot_type)) +
  geom_bar() +
  # angling axis text for readability
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 10
tennis_shots |> 
  ggplot(aes(x = shot_type)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::

## Behind the scenes of `geom_bar()`

-   Start with the data
-   Aggregate and count the number of observations in each bar
-   Map to plot aesthetics

![[R for data science (2e)](https://r4ds.hadley.nz/layers.html#the-layered-grammar-of-graphics)](https://r4ds.hadley.nz/images/visualization-grammar.png)

## Visualizing 1D categorical data

::: columns
::: {.column width="50%" style="text-align: left;"}

Instead of using `geom_bar()`, you can do these steps "by hand"

-   Aggregate and obtain the counts first with `count()` (or `group_by` and `summarize()`)

-   Then use `geom_col()`

```{r}
#| eval: false
tennis_shots |>
  count(shot_type, name = "count") |> 
  ggplot(aes(x = shot_type, y = count)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 10
tennis_shots |>
  count(shot_type, name = "count") |> 
  ggplot(aes(x = shot_type, y = count)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::

## Visualizing 1D categorical data

::: columns
::: {.column width="50%" style="text-align: left;"}
Flip your bar chart axes!

Simply replace `x` with `y`:

```{r}
#| eval: false
tennis_shots |> 
  ggplot(aes(y = shot_type)) +
  geom_bar() 
```

Or use `coord_flip()`:

```{r}
#| eval: false
tennis_shots |> 
  ggplot(aes(x = shot_type)) +
  geom_bar() +
  coord_flip() 
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig.height: 10
tennis_shots |> 
  ggplot(aes(x = shot_type)) +
  geom_bar() +
  coord_flip() 
```
:::
:::

## What does a bar chart show?

::: columns
::: {.column width="50%" style="text-align: left;"}

**Marginal distribution**: probability that a categorical variable $X$ (e.g., `shot_type`) takes each particular category value $x$ (`groundstroke`, `lob`, ..., `slice`)

::: {.incremental}

-   Frequency bar charts (earlier version) give info about sample size, but this could be labeled in the chart (use `geom_text()` or `geom_label()`)

-   Now, we create a proportion/percent bar chart to display the individual probabilities

-   This shows the probability mass function (PMF) for discrete variables

    -   For example: $P($ `shot_type` $=$ `slice`$)$
:::
:::

::: {.column width="50%" style="text-align: left;"}
:::{.fragment}
```{r}
#| fig-height: 9
tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n)) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col() # + geom_label(aes(label = n), hjust = 1)
```

:::
:::
:::


## Population vs sample

**Population:** The collection of all subjects of interest

**Sample:** A representative subset of the population of interest

. . .

<br>

*For us*: Our random sample of women's Wimbledon (non-serve) shots is a subset of all women's Wimbledon (non-serve) shots

. . .

<br>

**Empirical distribution**: estimates the true marginal distribution with observed (sample) data

. . .


:::{.incremental}
-   Estimate $P($ `shot_type` = $C_j$) with $\hat p_j$ for each category $C_j$ (e.g., $\hat p_\texttt{dropshot}$, ..., $\hat p_\texttt{volley}$)

 -   Standard error for each $\hat p_j$: $\quad \displaystyle \text{SE}(\hat{p}_j) = \sqrt{\frac{\hat{p}_j (1 - \hat{p}_j)}{n}}$
 
:::

## Adding confidence intervals to bar chart

::: columns
::: {.column width="50%" style="text-align: left;"}

We can add confidence intervals by calculating the standard error estimate for each category

```{r}
#| echo: true
#| eval: false
tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n),
         se = sqrt(prop * (1 - prop) / sum(n)),
         lower = prop - 2 * se,
         upper = prop + 2 * se) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col() +
  geom_errorbar(aes(xmin = lower, xmax = upper), 
                color = "blue", 
                width = 0.2, 
                linewidth = 1)
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 10
tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n),
         se = sqrt(prop * (1 - prop) / sum(n)),
         lower = prop - 2 * se,
         upper = prop + 2 * se) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col() +
  geom_errorbar(aes(xmin = lower, xmax = upper), 
                color = "blue", 
                width = 0.2, 
                linewidth = 1)
```
:::
:::

## Ordering factors in a bar chart

::: columns
::: {.column width="50%" style="text-align: left;"}

Let's order the bars by proportion

```{r}
#| eval: false
tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n),
         se = sqrt(prop * (1 - prop) / sum(n)),
         lower = prop - 2 * se,
         upper = prop + 2 * se,
         shot_type = fct_reorder(shot_type, prop)) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col() +
  geom_errorbar(aes(xmin = lower, xmax = upper), 
                color = "blue", width = 0.2, linewidth = 1)
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 10
tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n),
         se = sqrt(prop * (1 - prop) / sum(n)),
         lower = prop - 2 * se,
         upper = prop + 2 * se,
         shot_type = fct_reorder(shot_type, prop)) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col() +
  geom_errorbar(aes(xmin = lower, xmax = upper), 
                color = "blue", width = 0.2, linewidth = 1)
```
:::
:::

## Pie charts... don't make them

Why?

-   <https://www.data-to-viz.com/caveat/pie.html>

-   <https://github.com/cxli233/FriendsDontLetFriends>

. . .

*A particularly painful pie chart from Fox News on 2012 GOP candidates:*

::: {fig-align="center"}
![Courtesy of [dataviz.ca](https://www.datavis.ca/gallery/evil-pies.php)](https://www.datavis.ca/gallery/images/pies/FoxNews-GOP.png){width="40%"}
:::

. . . 

3D pie charts?... even worse

## Inference for 1D categorical data

**Chi-square test for 1D categorical data**

-   Null hypothesis: $H_0$: $p_1 = p_2 = \cdots = p_K$

-   Alternative hypothesis: $H_1$: At least one category has a different proportion

. . .

*Test statistic:*

$$\displaystyle \chi^2 = \sum_{j=1}^K \frac{(O_j - E_j)^2}{E_j},$$

where

  -   $O_j$: observed counts in category $j$

   -   $E_j$ : expected counts under the null (i.e., $n/K$ or each category is equally likely to occur)

. . .

*We can perform this test in `R`*:

```{r}
#| eval: false
chisq.test(table(tennis_shots$shot_type))
```

## Hypothesis testing in general

Computing $p$-values works like this:

:::{.incremental}

-   Choose a test statistic

-   Compute the test statistic using the observed data

-   Is test statistic "unusual" compared to what we would expect under the null hypothesis ($H_0$)?

-   Compare $p$-value to the pre-specified target Type I error rate ("significance level") $\alpha$

    -   Typically choose $\alpha = 0.05$ ([the origins of 0.05](https://www2.psych.ubc.ca/~schaller/528Readings/CowlesDavis1982.pdf))

:::{.fragment}

<br>

  
```{r}
#| echo: false
library(gt)
library(katex)
library(knitr)
library(kableExtra)

tribble(
  ~Decision, ~`H0 true`, ~`H1 true`,
  "Fail to reject \\(H_0\\)", "Correct decision", "Type II error (\\(\\beta\\))",
  "Reject \\(H_0\\)", "Type I error (\\(\\alpha\\))", "Correct decision") |>
  kable(
    format = "html",
    escape = FALSE,
    col.names = c("Decision", "\\(H_0\\) true", "\\(H_1\\) true")
  )
```
:::
:::

# 2D categorical data

## Summarizing 2D categorical data


Continuing with our women's Wimbledon tennis sample, let's look at two attributes:

-   `shot_type` (What was the shot type?)

-   `outcome_type` (What was the shot's outcome?)


. . .

*How many levels does each variable have?*

```{r}
table(tennis_shots$shot_type)
table(tennis_shots$outcome_type)
```

. . .

Let's put the two error categories in `outcome_type` together...

```{r}
tennis_shots <- mutate(tennis_shots, outcome_type = fct_relevel(outcome_type, "unforced error"))
```


## Summarizing 2D categorical data

**Two-way table** (or contingency table, cross tabulation, crosstab)

```{r}
table("Shot type" = tennis_shots$shot_type, 
      "Outcome type" = tennis_shots$outcome_type)

xtabs(~ shot_type + outcome_type, data = tennis_shots)
```

## Visualizing 2D categorical data

::: columns
::: {.column width="50%" style="text-align: left;"}

**Stacked bar chart:** a bar chart of spine charts

Emphasizes the marginal distribution of each category of the `X` variable

-   e.g., $P($ `outcome_type` $=$ `winner` $)$

:::{.fragment}

*Similar to 1D bar charts:* 

1. Start with counting every combination of 2 variables (using `count()` or `group_by()` and `summarize()`),

2. Plot with `geom_col()`

```{r}
#| eval: false
#| echo: false
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n, 
             # filled by the other categorical variable
             fill = outcome_type)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::

::: {.column width="50%" style="text-align: left;"}
:::{.fragment}
```{r}
#| echo: true
#| eval: true
#| fig-height: 9
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n, fill = outcome_type)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::
:::

## Visualizing 2D categorical data

::: columns
::: {.column width="50%" style="text-align: left;"}

**Stacked bar chart** (proportion version)

```{r}
#| eval: false
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n,
             fill = outcome_type)) +
  geom_col(position = "fill") +
  labs(y = "prop") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```

*Just make sure to change the y-axis title (at least)!*
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 10
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n,
             fill = outcome_type)) +
  geom_col(position = "fill") +
  labs(y = "prop") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::

## Visualizing 2D categorical data

::: columns
::: {.column width="50%" style="text-align: left;"}
Side-by-side (grouped, dodged) bar chart: a bar chart of bar charts

Shows the conditional distribution of `fill` variable given `X` variable

*For example:* $P($ `outcome_type` $=$ `winner` $\mid$ `shot_type` $=$ `groundstroke` $)$

```{r}
#| eval: false
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n,
             fill = outcome_type)) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 10
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n,
             fill = outcome_type)) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::

## Joint, marginal, and conditional probabilities

Let $X$ = `shot_type` and $Y$ = `outcome_type`

:::{.incremental}

-   **Joint distribution:** frequency of the intersection

    -   e.g., $P(X =$ `overhead` $, Y =$ `winner` $)$

-   **Marginal distribution:** row sums or column sums

    -   e.g., $P(X =$ `overhead` $)$, $P(Y =$ `winner` $)$

-   **Conditional distribution:** probability event $X$ given event $Y$

    -   e.g., $P(X =$ `overhead` $\mid Y =$ `winner` $)$ <br><br>$\displaystyle \qquad \quad = \frac{P(X = \texttt{overhead}, Y = \texttt{winner})}{P(Y = \texttt{winner})}$
:::



## Joint, marginal, and conditional probabilities

**Two-way proportion table** (the `tidyverse` way) with `pivot_wider`

```{r}
tennis_shots |>
  group_by(shot_type, outcome_type) |>
  summarize(joint = n() / nrow(tennis_shots)) |>
  pivot_wider(names_from = outcome_type, values_from = joint, values_fill = 0)
```

<br>

```{r}
#| eval: false
# The base `R` way to get a two-way proportion table
tennis_shots |> 
  select(shot_type, outcome_type) |> 
  table() |> 
  prop.table()
```

## Categorical heatmaps

::: columns
::: {.column width="50%" style="text-align: left;"}
-   Use `geom_tile()` to display joint distribution of two categorical variables

-   Annotate tiles with labels of percentages using `geom_text()` and the `scales` package (a very neat package)

```{r}
#| eval: false

#install.packages("scales")

tennis_shots |>
  group_by(shot_type, outcome_type) |>
  summarize(
    freq = n(), 
    joint = n() / nrow(tennis_shots)
  ) |> 
  ggplot(aes(x = outcome_type, y = shot_type)) +
  geom_tile(aes(fill = freq), color = "white") +
  geom_text(aes(label = scales::percent(joint, accuracy = 0.01))) +
  scale_fill_gradient2() +
  coord_equal() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 11
tennis_shots |>
  group_by(shot_type, outcome_type) |>
  summarize(
    freq = n(), 
    joint = n() / nrow(tennis_shots)
  ) |> 
  ggplot(aes(x = outcome_type, y = shot_type)) +
  geom_tile(aes(fill = freq), color = "white") +
  geom_text(aes(label = scales::percent(joint, accuracy = 0.01))) +
  scale_fill_gradient2() +
  coord_equal() +
  theme_classic() +
  theme(text = element_text(size = 20),
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
:::

## Visualizing independence

::: columns
::: {.column width="50%" style="text-align: left;"}

**Mosaic plot**

-   A spine chart of spine charts

-   *Width:* marginal distribution of `shot_type`

-   *Height:* conditional distribution of `outcome_type` \| `shot_type`

-   *Area:* joint distribution

:::{.fragment}

Using a mosaic plot to visually check for independence:

-   Check whether all proportions are the same (i.e., the boxes line up in a grid)
:::
:::{.fragment}

```{r}
#| fig-height: 10
#| echo: true
#| eval: false
tennis_shots |> 
  select(shot_type, outcome_type) |> 
  table() |> 
  mosaicplot(main = "Relationship between shot type and outcome type",
             las=2, cex.axis = 1.2, cex.lab = 1.5, cex.main = 2)
```
:::
:::

::: {.column width="50%" style="text-align: left;"}
:::{.fragment}
```{r}
#| fig-height: 12
#| echo: false
tennis_shots |> 
  select(shot_type, outcome_type) |> 
  table() |> 
  mosaicplot(main = "Relationship between shot type and outcome type",
             las=2, cex.axis = 1.5, cex.lab = 2, cex.main = 2)
```
:::
:::
:::

## Visualizing independence




Mosaic plot with `ggmosaic` package


:::{.columns}

::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: true
#| eval: false
# devtools::install_github("haleyjeppson/ggmosaic")
library(ggmosaic)
tennis_shots |> 
  ggplot() +
  geom_mosaic(aes(x = product(outcome_type, shot_type),
                  fill = outcome_type)) +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::
::: {.column width="50%" style="text-align: left;"}
```{r}
#| echo: false
#| fig-height: 10
#| fig-align: center
# devtools::install_github("haleyjeppson/ggmosaic")
library(ggmosaic)
tennis_shots |> 
  ggplot() +
  geom_mosaic(aes(x = product(outcome_type, shot_type),
                  fill = outcome_type)) +
  theme_classic() +
  theme(text = element_text(size = 20),
        legend.position = "top",
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```
:::

:::
::: aside
*See also:* [Quang Nguyen's blog on making mosaic plots from scratch with `ggplot2`](https://qntkhvn.netlify.app/posts/2026-05-17-mosaic-ggplot2/).
:::

## Inference for 2D categorical data

**Chi-square test for 2D categorical data**

-   Null hypothesis: $H_0$: two categorical variables are independent of each other

    -   e.g., no association between `shot_type` and `outcome_type`

- Alternative hypothesis: $H_1$: two categorical variables are dependent

. . .

*Test statistic:*

$$\displaystyle \chi^2 = \sum_i^{k_1} \sum_j^{k_2} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

. . .

*We can alslo perform this test in `R`*:

```{r}
#| eval: false
# chisq.test(table(tennis_shots$outcome_type, tennis_shots$shot_type))

tennis_shots |> 
  select(outcome_type, shot_type) |> 
  table() |> 
  chisq.test()
```

## Residuals

Recall the test statistic:

$$\displaystyle \chi^2 = \sum_i^{k_1} \sum_j^{k_2} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

Define the Pearson residuals:

$$\displaystyle r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}}}$$

. . .

Some rules of thumb:

-   $r_{ij} \approx 0$: observed counts are close to expected counts

-   $|r_{ij}| > 2$: significant at $\alpha = 0.05$

-   very positive $r_{ij}$: higher than expected

-   very negative $r_{ij}$: lower than expected

## Residuals

:::{.columns}
::: {.column width="50%" style="text-align: left;"}

- We can make mosaic plots with the cells color-coded by Pearson residuals

- Tells us which combinations of 2 categorical variables (cells) are much higher/lower than expected

```{r}
#| eval: false
tennis_shots |> 
  select(shot_type, outcome_type) |> 
  table() |> 
  mosaicplot(main = "Relationship between outcome type and shot type",
             shade = TRUE, las = 2,
             cex.axis = 1.2, cex.lab = 2, cex.main = 2)
```
:::
::: {.column width="50%" style="text-align: left;"}
```{r}
#| fig-height: 12
#| fig-align: center
#| echo: false
tennis_shots |> 
  select(shot_type, outcome_type) |> 
  table() |> 
  mosaicplot(main = "Relationship between outcome type and shot type",
             shade = TRUE, las = 2,
             cex.axis = 1.2, cex.lab = 2, cex.main = 2)
```
:::

:::

## Beyond 2D: facets!

*Using some* `dplyr` *data wrangling and* `ggplot2` *material...*

```{r}
#| fig-align: "center"
#| fig-width: 20
tennis_shots |>
  filter(!(round %in% c("Q1", "Q2", "Q3"))) |>
  mutate(round_grouped = case_when(
    round == "F" ~ "Finals",
    round == "SF" ~ "Semi-finals",
    round == "QF" ~ "Quarter-finals",
    TRUE ~ "Early rounds")) |>
  mutate(round_grouped = fct_relevel(round_grouped, "Early rounds", "Quarter-finals", "Semi-finals", "Finals")) |>
  ggplot(aes(x = shot_type, fill = outcome_type)) + 
  geom_bar(position = "fill") + # could remove position = "fill" to get raw counts
  labs(x = "Shot type", y = "Proportion", fill = "Outcome Type") +
  facet_wrap(~ round_grouped, ncol = 4) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# Appendix

## The `janitor` package

The most popular `janitor` function is `clean_names()`... for cleaning column names

```{r}
# before
iris |> 
  head()

# after
library(janitor)
iris |> 
  clean_names() |> 
  head()
```

## Tabulation with the `janitor` package

The lesser-known stars of `janitor`: functions for tabulation of categorical data


The `tabyl()` function:

```{r}
library(gt) # for nicely outputted tables

tennis_shots |> 
  tabyl(shot_type) |>
  gt()
```

:::{.aside}
For more, see [this overview](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html) and [this tutorial](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html).
:::

## Tabulation with the `janitor` package

The `adorn_*()` functions

```{r}
tennis_shots |> 
  tabyl(shot_type, outcome_type) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting(digits = 2) |> 
  adorn_ns() |>
  gt()
```

<!-- ::: columns -->

<!-- ::: {.column width="50%" style="text-align: left;"} -->

<!-- c1 -->

<!-- ::: -->

<!-- ::: {.column width="50%" style="text-align: left;"} -->

<!-- c2 -->

<!-- ::: -->

<!-- ::: -->
