---
title: "Unsupervised learning: $k$-means clustering"
author: "<br>CMSACamp 2026<br><br>Department of Statistics & Data Science<br>Carnegie Mellon University"
footer:  "[36-SURE.github.io/2026](https://36-sure.github.io/2026)"
format:
  revealjs:
    theme: theme.scss
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.5
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Background

## Into statistical learning with unsupervised learning

What is **statistical learning**?

> Statistical learning refers to a set of tools for modeling and understanding complex datasets.--- Preface of Introduction to Statistical Learning with Applications in R (ISLR) 

. . .

In **unsupervised learning**, we have a set of features but there are no *response* variables (i.e., the data is not "labeled")

. . .

<br>

We can think of unsupervised learning as *an extension of EDA*

- Often, the goal is to identify patterns, subgroups, and/or relationships in the data

- Like EDA, there is no unique right answer in unsupervised learning

. . .


## Clustering (cluster analysis)

What is **clustering** (aka **cluster analysis**)?

>  Clustering refers to a very broad set of techniques for finding subgroups, or clusters, in a data set --- Section 10.3 of ISLR

. . .

*In particular:*

-   Observations **within** clusters are **more similar** to each other

-   Observations **in different** clusters are **more different** from each other

. . .

*Notes:*

- Clustering often involves domain-specific considerations based on knowledge of the data

- Clustering methods can be applied to anything that can be represented as tabular data

  - E.g., trajectory curves, shapes, graphs/networks, and images 


## Types of clustering

There are two main types of clustering:

- **Hard clustering** ($k$-means, hierarchical)

  - Output "hard" assignment
  
  - Strictly assign observations to only one cluster
  
  
. . .


- **Soft/fuzzy clustering** (model-based clustering, mixture models)


  - Output "soft" assignments
  
  - Assign an observation a probability of belonging to each cluster
  
  - Provide uncertainty in the clustering results
  
  - Incorporate statistical modeling into clustering


## Euclidean distance

Suppose we have two observations: $\boldsymbol{x}_i = (x_{i1}, \dots, x_{ip})$ and $\boldsymbol{x}_j = (x_{j1}, \dots, x_{jp})$

. . .

<br>

How do we define **distance/dissimilarity** between these observations?

. . .

<br>

A common way to measure the distance is with the **Euclidean distance**: 


$$d(\boldsymbol{x}_i, \boldsymbol{x}_j) = \sqrt{(x_{i1} - x_{j1})^2 + \dots + (x_{ip} - x_{jp})^2}$$

. . .

<br>

*Important: scaling of the variables matters!*

- One variable may *dominate* others when computing Euclidean distance because its range is much larger

- We will standardize each variable to have mean 0 and standard deviation 1 with `scale()`

---

## Distances in general

* There's a variety of different types of distance metrics: [Manhattan](https://en.wikipedia.org/wiki/Taxicab_geometry), [Mahalanobis](https://en.wikipedia.org/wiki/Mahalanobis_distance), [Cosine](https://en.wikipedia.org/wiki/Cosine_similarity), [Kullback-Leibler](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence), [Hellinger](https://en.wikipedia.org/wiki/Hellinger_distance),
[Wasserstein](https://en.wikipedia.org/wiki/Wasserstein_metric)

* We're just going to focus on [Euclidean distance](https://en.wikipedia.org/wiki/Euclidean_distance)

. . .

We can represent the collection of pairwise distances between observations $\boldsymbol{x}_1, \dots, \boldsymbol{x}_n$ with a **distance matrix**: 

$$
D = \begin{pmatrix}
                0 & D_{12} & \cdots & D_{1n} \\
                D_{21} & 0 & \cdots & D_{2n} \\
                \vdots & \vdots & \ddots & \vdots \\
                D_{n1} & \cdots & \cdots & 0
            \end{pmatrix},
$$
<br>

where: $D_{ij} = d(\boldsymbol{x}_i, \boldsymbol{x}_j)$, $D_{ij} = D_{ji}$, and $D_{ii} = 0$

# $k$-means clustering

## $k$-means clustering

**Goal**: partition the observations into a pre-specified number of clusters by minimizing the total *within-cluster variation*

. . .

<br>

Let $C_1, \dots, C_K$ denote sets containing indices of observations in each of the $K$ clusters

- If observation $i$ is in cluster $k$, then $i \in C_k$

. . .

Define the total *within-cluster variation* as the sum of the within-cluster variation $W(C_k)$ for each cluster $k$

- Therefore, for $k$-means clustering, we solve: 


$$
\displaystyle \underset{C_1, \dots, C_K}{\text{minimize }} \Big\{ \sum_{k=1}^K W(C_k) \Big\}
$$


## $k$-means clustering

How do we define *within-cluster variation* for a cluster $k$?

. . .

<br>

We can use the **(squared) Euclidean distance**:

$$
W(C_k) = \frac{1}{|C_k|}\sum_{i,j \in C_k} d(x_i, x_j)^2 \,,
$$

where $|C_k|$ denotes the number of observations in cluster $k$

. . .

<br>

This is commonly referred to as the **within-cluster sum of squares (WSS)**

. . .

<br>

::: {style="text-align: center;"}
**So, how can we solve this?**
:::

## Lloyd's algorithm (naive $k$-means)

1)  Choose $K$ random centers, aka **centroids** (hence the name $k$-means)

2) Assign each observation closest center (using Euclidean distance)

3) Repeat until cluster assignment stop changing:

  - Compute new centroids as the averages of the updated groups
  
  - Reassign each observations to closest center

. . .
  

::: {style="text-align: center;"}
[Visualization of Lloyd's algorithm](https://www.youtube.com/watch?v=5I3Ei69I40s)
:::

. . .

*Some notes:*

- **Typically converges to a local optimum**; we're not guaranteed to find the global optimum of the objective function (WSS)

- **Results will change from run to run** ([see this video](https://www.youtube.com/watch?v=9nKfViAfajY))

- **Takes $K$ as an input!**

## Division 1 volleyball data

Each row represents a team at the NCAA Division 1 level from the 2022-2023 season

:::{.aside}
The data is originally from the [SCORE Sports Data Repository](https://data.scorenetwork.org/volleyball/volleyball_ncaa_team_stats.html).
:::

```{r}
#| echo: false
set.seed(47)
```

```{r}
library(tidyverse)
theme_set(theme_light())

d1_volleyball <- read_csv("https://data.scorenetwork.org/data/volleyball_ncaa_div1_2022_23.csv")

glimpse(d1_volleyball)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```

## $k$-means clustering example 

::: columns
::: {.column width="50%" style="text-align: left;"}

We can do k-means clustering via the `kmeans()` function in `R`, but we **must** provide number of clusters (here we set $K = 4$)

```{r}
#| label: first-kmeans
#| eval: false
init_kmeans <- d1_volleyball |> 
  select(kills_per_set, blocks_per_set) |> 
  kmeans(algorithm = "Lloyd", centers = 4,
         nstart = 1, iter.max = 20)

d1_volleyball |>
  mutate(
    team_clusters = as.factor(init_kmeans$cluster)
  ) |>
  ggplot(aes(x = blocks_per_set, y = kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom")
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: first-kmeans
#| echo: false
#| fig-height: 9
```
:::
:::


*This example only uses 2 features (`blocks_per_set` and `kills_per_set`), but in practice, you can (and should) include more than two features*

## Careful with units...

::: columns
::: {.column width="50%" style="text-align: left;"}

- We can use `coord_fixed()` so that the axes match with unit scales

- `blocks_per_set` and `kills_per_set` are on very different scales

```{r}
#| label: coord-fixed
#| eval: false
d1_volleyball |>
  mutate(
    team_clusters = as.factor(init_kmeans$cluster)
  ) |>
  ggplot(aes(x = blocks_per_set, y = kills_per_set, 
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom") +
  coord_fixed()
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: coord-fixed
#| echo: false
#| fig-height: 12
```
:::
:::

## Standardize the variables!

::: columns
::: {.column width="50%" style="text-align: left;"}

We use the `scale()` function to first **standardize** the variables:

$$
\displaystyle \frac{\text{value} - \text{mean}}{\text{standard deviation}}
$$

```{r}
#| label: std-kmeans
#| eval: false
clean_d1_volleyball <- d1_volleyball |>
  mutate(
    std_kills_per_set = as.numeric(scale(kills_per_set, center = TRUE, scale = TRUE)),
    std_blocks_per_set = as.numeric(scale(blocks_per_set, center = TRUE, scale = TRUE))
  )

std_kmeans <- clean_d1_volleyball |> 
  select(std_kills_per_set, std_blocks_per_set) |> 
  kmeans(algorithm = "Lloyd", centers = 4, nstart = 1, iter.max = 20)

clean_d1_volleyball |>
  mutate(
    team_clusters = as.factor(std_kmeans$cluster)
  ) |>
  ggplot(aes(x = blocks_per_set, y = kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom") +
  coord_fixed()
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: std-kmeans
#| echo: false
#| fig-height: 12
```
:::
:::

## Standardize the variables!

::: columns
::: {.column width="50%" style="text-align: left;"}

We can also plot the standardized average kills per set and standardized blocks per set


```{r}
#| label: std-kmeans-view
#| eval: false
clean_d1_volleyball |>
  mutate(
    team_clusters = as.factor(std_kmeans$cluster)
  ) |>
  ggplot(aes(x = std_blocks_per_set, y = std_kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom") +
  coord_fixed()
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: std-kmeans-view
#| echo: false
#| fig-height: 12
```
:::
:::

## And if we run $k$-means again?

::: columns
::: {.column width="50%" style="text-align: left;"}
We get different clustering results!

```{r}
#| label: second-kmeans
#| eval: false
another_kmeans <- clean_d1_volleyball |> 
  select(std_kills_per_set, std_blocks_per_set) |> 
  kmeans(algorithm = "Lloyd", centers = 4, nstart = 1)

clean_d1_volleyball |>
  mutate(
    team_clusters = as.factor(another_kmeans$cluster)
  ) |>
  ggplot(aes(x = blocks_per_set, y = kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom")
```

**Results depend on initial centroids**

*Keep in mind: the labels / colors are arbitrary*
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: second-kmeans
#| echo: false
#| fig-height: 12
```
:::
:::

## Fix randomness issue with `nstart`

::: columns
::: {.column width="50%" style="text-align: left;"}

Run the algorithm `nstart` times, then pick the results with *lowest total within-cluster variation*: $$\text{total WSS} = \sum_{k=1}^K W(C_k)$$

```{r}
#| label: nstart-kmeans
#| eval: false
nstart_kmeans <- clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  kmeans(algorithm = "Lloyd", centers = 4,
         nstart = 30, iter.max = 20)

clean_d1_volleyball |>
  mutate(
    team_clusters = as.factor(nstart_kmeans$cluster)
  ) |> 
  ggplot(aes(x = blocks_per_set, y = kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom")
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: nstart-kmeans
#| echo: false
#| fig-height: 12
```
:::
:::

## By default `R` uses [Hartigan–Wong method](https://en.wikipedia.org/wiki/K-means_clustering#Hartigan%E2%80%93Wong_method)

::: columns
::: {.column width="50%" style="text-align: left;"}

- Updates based on changing the cluster assignment of a single observation

- Computational advantages over re-computing distances for every observation

```{r}
#| label: default-kmeans
#| eval: false

default_kmeans <- clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  kmeans(algorithm = "Hartigan-Wong", centers = 4,
         nstart = 30, iter.max = 20) 

clean_d1_volleyball |>
  mutate(
    team_clusters = as.factor(default_kmeans$cluster)
  ) |> 
  ggplot(aes(x = blocks_per_set, y = kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom")
```

*Very little difference for our purposes...*
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: default-kmeans
#| echo: false
#| fig-height: 12
```
:::
:::

## Clustering with more than 2 variables

1. First, we select the variables

```{r}
d1_volleyball_features <- d1_volleyball |>
  select(blocks_per_set, digs_per_set, kills_per_set, aces_per_set) |> 
  drop_na() 
```

2. Next, standardize the variables (as always)

```{r}
std_d1_volleyball_features <- d1_volleyball_features |> 
  scale(center = TRUE, scale = TRUE)
```

3. Now, perform clustering

```{r}
kmeans_many_features <- std_d1_volleyball_features |> 
  kmeans(algorithm = "Hartigan-Wong", centers = 4, nstart = 30) 
```

## Clustering with more than 2 variables

One way to examine the clusters is by looking at the (non-standardized) means for each variable included in the cluster analysis

```{r}
library(gt)

clean_d1_volleyball |>
  select(blocks_per_set, digs_per_set, kills_per_set, aces_per_set) |> 
  mutate(team_clusters = as.factor(kmeans_many_features$cluster)) |> 
  pivot_longer(-team_clusters, names_to = "feature", values_to = "value") |>
  group_by(team_clusters, feature) |>
  summarize(avg_value = mean(value), n_teams = n(), .groups = "drop") |>
  pivot_wider(id_cols = c(team_clusters, n_teams), names_from = feature, values_from = avg_value) |>
  gt() |>
  fmt_number(columns = -n_teams, decimals = 2)
  
```

We can also do more EDA: *e.g., do teams in the same conference tend be in the same cluster? How does cluster membership relate to win/loss percentage?*



## Beyond $k$-means: $k$-means++

**Goal**: initialize the cluster centers before proceeding with the standard $k$-means clustering algorithm, and provide a better alternative to `nstart`

. . .

<br>

*At a high-level*: 

* Randomly select an observation to be the first cluster center

* Each subsequent cluster center is chosen from the remaining observations

  * The probability that an observation chosen as a cluster center is proportional to its squared distance from the closest existing cluster center

. . .

<br>

::: {style="text-align: center;"}
[Visualization of $k$-means++](https://www.youtube.com/watch?v=BIQDlmZDuf8)
:::

## The $k$-means++ algorithm

1. Pick a random observation to be the center $c_1$ of the first cluster $C_1$

    -   This initializes a set of centers $\mathscr C = \{c_1 \}$

. . .

2. For each observation (that is not a center), compute $D(x_i) = \underset{c \in \mathscr C}{\text{min}} \ d(x_i, c)$

    -   Distance between observation and its closest center $c \in \mathscr C$

. . .

3. Randomly pick a point $x_i$ with probability: $\displaystyle p_i = \frac{D^2(x_i)}{\sum_{j=1}^n D^2(x_j)}$

    - As distance to closest center increases, the probability of selection increases
  
. . .

4. Update $\mathscr C = \mathscr C \cup c^*$

. . .

5. Repeat steps 3 and 4 until there are $K$ centers

. . .

6. Run $k$-means using these $\mathscr C$ as the starting points

. . .




## $k$-means++ in `R` using [`flexclust`](https://cran.r-project.org/web/packages/flexclust/flexclust.pdf)

::: columns
::: {.column width="50%" style="text-align: left;"}
```{r}
#| label: kmeanspp
#| eval: false
library(flexclust)
init_kmeanspp <- clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  kcca(k = 4, control = list(initcent = "kmeanspp"))

clean_d1_volleyball |>
  mutate(
    team_clusters = as.factor(init_kmeanspp@cluster)
  ) |>
  ggplot(aes(x = blocks_per_set, y = kills_per_set,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom")
```

*Note the use of `@` instead of `$`...*
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| ref-label: kmeanspp
#| echo: false
#| fig-height: 12
```
:::
:::

## So, how do we choose the number of clusters?

**There is no universally accepted way to conclude that a particular choice of $k$ is optimal!**

. . .

From [Cosma Shalizi's notes](https://www.stat.cmu.edu/~cshalizi/350/lectures/08/lecture-08.pdf)

> One reason you should be intensely skeptical of clustering results --- including your own! --- is that there is currently very little theory about how to find the right number of clusters. It’s not even completely clear what "the right number of clusters" means!

Additional reading: [Statistical Methods for Machine Learning notes](https://www.stat.cmu.edu/~larry/=sml/clustering.pdf)

## Popular heuristic: elbow plot (use with caution)

Looks at the total within-cluster variation as a function of the number of clusters

*Let's do this by hand first...*

::: columns
::: {.column width="50%" style="text-align: left;"}
```{r}
# function to perform clustering for each value of k
d1_volleyball_kmeans <- function(k) {
  
  kmeans_results <- clean_d1_volleyball |>
    select(std_blocks_per_set, std_kills_per_set) |>
    kmeans(centers = k, nstart = 30)
  
  kmeans_out <- tibble(
    clusters = k,
    total_wss = kmeans_results$tot.withinss
  )
  return(kmeans_out)
}
```
:::

::: {.column width="50%" style="text-align: left;"}
```{r}
#| eval: false
# number of clusters to search over
n_clusters_search <- 2:15

# iterate over each k to compute total wss
kmeans_search <- n_clusters_search |> 
  map(d1_volleyball_kmeans) |> 
  bind_rows()

# plot the results
kmeans_search |> 
  ggplot(aes(x = clusters, y = total_wss)) +
  geom_line() + 
  geom_point(size = 4) +
  scale_x_continuous(breaks = n_clusters_search)
```
:::
:::

## Popular heuristic: elbow plot (use with caution)

::: columns
::: {.column width="50%" style="text-align: left;"}

- Choose $k$ where marginal improvements is low at the bend (hence the elbow)

  - Based on the plot, this would somewhere around 5--8 clusters

- **This is just a guideline and should not dictate your choice of** $k$

- Other choices: [gap statistic](https://web.stanford.edu/~hastie/Papers/gap.pdf), [silhouette](https://en.wikipedia.org/wiki/Silhouette_(clustering))

:::

::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 10
# number of clusters to search over
n_clusters_search <- 2:15

# iterate over each k to compute total wss
kmeans_search <- n_clusters_search |> 
  map(d1_volleyball_kmeans) |> 
  bind_rows()

kmeans_search |> 
  ggplot(aes(x = clusters, y = total_wss)) +
  geom_line() + 
  geom_point(size = 4) +
  scale_x_continuous(breaks = n_clusters_search)
```

:::
:::

## Appendix: elbow method with `factoextra`

::: columns
::: {.column width="50%" style="text-align: left;"}

We can also make elbow plots with the `factoextra` package (*also based on total WSS*)

```{r}
#| echo: true
#| eval: false

library(factoextra)

clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  fviz_nbclust(kmeans, method = "wss")
```

:::

::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 9
library(factoextra)

clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  fviz_nbclust(kmeans, method = "wss")
```

:::

:::

## Appendix: silhouette method with `factoextra`

::: columns
::: {.column width="50%" style="text-align: left;"}

We can also make a silhouette plot with `factoextra`

```{r}
#| echo: true
#| eval: false
clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  fviz_nbclust(kmeans, method = "silhouette")
```

*For more details, see: [silhouette (clustering)](https://en.wikipedia.org/wiki/Silhouette_(clustering))*

:::

::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 12
clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  fviz_nbclust(kmeans, method = "silhouette")
```

:::

:::

## Appendix: gap statistic

::: columns
::: {.column width="50%" style="text-align: left;"}

We could also compute the gap statistics using the `cluster` package

```{r}
#| echo: true
library(cluster)

d1_volleyball_kmeans_gap_stat <- clean_d1_volleyball |> 
  select(std_blocks_per_set, std_kills_per_set) |> 
  clusGap(FUN = kmeans, nstart = 30, K.max = 10)

# view the result 
#d1_volleyball_kmeans_gap_stat |> 
#  print(method = "firstmax")
```

*For more details, see: [gap statistic](https://web.stanford.edu/~hastie/Papers/gap.pdf)*

:::{.fragment}

<br>

**Notice all these methods give a different number of "optimal" clusters!**

:::

:::

::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: true
#| fig-height: 10
d1_volleyball_kmeans_gap_stat |> 
  fviz_gap_stat(maxSE = list(method = "firstmax"))
```

:::

:::

## Appendix: elbow plot for $k$-means++

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r}
#| label: kmeanspp-elbow
#| eval: false
d1_volleyball_kmpp <- function(k) {
  
  kmeans_results <- clean_d1_volleyball |>
    select(std_blocks_per_set, std_kills_per_set) |>
    kcca(k = k, control = list(initcent = "kmeanspp"))
  
  kmeans_out <- tibble(
    clusters = k,
    total_wss = sum(kmeans_results@clusinfo$size * 
                      kmeans_results@clusinfo$av_dist)
  )
  return(kmeans_out)
}

n_clusters_search <- 2:12

kmpp_search <- n_clusters_search |> 
  map(d1_volleyball_kmpp) |> 
  bind_rows()

kmpp_search |> 
  ggplot(aes(x = clusters, y = total_wss)) +
  geom_line() + 
  geom_point(size = 2) +
  scale_x_continuous(breaks = n_clusters_search)
```
:::

::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 11
d1_volleyball_kmpp <- function(k) {
  
  kmeans_results <- clean_d1_volleyball |>
    select(std_blocks_per_set, std_kills_per_set) |>
    kcca(k = k, control = list(initcent = "kmeanspp"))
  
  kmeans_out <- tibble(
    clusters = k,
    total_wss = sum(kmeans_results@clusinfo$size * 
                      kmeans_results@clusinfo$av_dist)
  )
  return(kmeans_out)
}

n_clusters_search <- 2:12

kmpp_search <- n_clusters_search |> 
  map(d1_volleyball_kmpp) |> 
  bind_rows()

kmpp_search |> 
  ggplot(aes(x = clusters, y = total_wss)) +
  geom_line() + 
  geom_point(size = 2) +
  scale_x_continuous(breaks = n_clusters_search)
```

:::
:::



<!-- ## Appendix: $k$-means for image segmentation and compression -->

<!-- Goal: partition an image into multiple segments, where each segment typically represents an object in the image -->

<!-- * Treat each pixel in the image as a point in 3-dimensional space comprising the intensities of the (red, blue, green) channels -->

<!-- * Treat each pixel in the image as a separate data point -->

<!-- *  Apply $k$-means clustering and identify the clusters -->

<!-- *  All the pixels belonging to a cluster are treated as a segment in the image -->

<!-- Then for any $k$, reconstruct the image by replacing each pixel vector with the (red, blue, green) triplet given by the center to which that pixel has been assigned -->

<!-- ## Appendix: $k$-means for image segmentation and compression -->

<!-- ```{r} -->
<!-- # https://raw.githubusercontent.com/36-SURE/2024/main/data/spongebob.jpeg -->
<!-- set.seed(2) -->
<!-- library(jpeg) -->
<!-- img_raw <- readJPEG("../data/spongebob.jpeg") -->
<!-- img_width <- dim(img_raw)[1] -->
<!-- img_height <- dim(img_raw)[2] -->
<!-- img_tbl <- tibble(x = rep(1:img_height, each = img_width), -->
<!--                   y = rep(img_width:1, img_height), -->
<!--                   r = as.vector(img_raw[, , 1]), -->
<!--                   g = as.vector(img_raw[, , 2]), -->
<!--                   b = as.vector(img_raw[, , 3])) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- img_kmpp <- function(k) { -->
<!--   km_fit <- img_tbl |> -->
<!--     select(r, g, b) |> -->
<!--     kcca(k = k, control = list(initcent = "kmeanspp")) -->
<!--   km_col <- rgb(km_fit@centers[km_fit@cluster,]) -->
<!--   return(km_col) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- img_tbl <- img_tbl |> -->
<!--   mutate(original = rgb(r, g, b), -->
<!--          k2 = img_kmpp(2), -->
<!--          k4 = img_kmpp(4), -->
<!--          k8 = img_kmpp(8), -->
<!--          k11 = img_kmpp(11)) |> -->
<!--   pivot_longer(original:k11, names_to = "k", values_to = "hex") |> -->
<!--   mutate(k = factor(k, levels = c("original", "k2", "k4", "k8", "k11"))) -->
<!-- ``` -->

<!-- ## Appendix: $k$-means for image segmentation and compression -->




<!-- ```{r} -->
<!-- #| eval: false -->
<!-- img_tbl |>  -->
<!--   ggplot(aes(x, y, color = hex)) + -->
<!--   geom_point() + -->
<!--   scale_color_identity() + -->
<!--   facet_wrap(~ k, nrow = 1) + -->
<!--   coord_fixed() + -->
<!--   theme_classic() -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #| echo: false -->
<!-- img_tbl |>  -->
<!--   ggplot(aes(x, y, color = hex)) + -->
<!--   geom_point() + -->
<!--   scale_color_identity() + -->
<!--   facet_wrap(~ k, nrow = 1) + -->
<!--   coord_fixed() + -->
<!--   theme_void() + -->
<!--   theme(strip.text = element_text(size = 14)) -->
<!-- ``` -->


<!-- ::: columns -->

<!-- ::: {.column width="50%" style="text-align: left;"} -->

<!-- c1 -->

<!-- ::: -->

<!-- ::: {.column width="50%" style="text-align: left;"} -->

<!-- c2 -->

<!-- ::: -->

<!-- ::: -->
