---
title: "Presentations"
author: "<br>SURE 2025<br><br>Department of Statistics & Data Science<br>Carnegie Mellon University"
footer:  "[36-SURE.github.io/2025](https://36-sure.github.io/2025)"
format:
  revealjs:
    theme: theme.scss
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.25
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Timeline

## EDA project presentation

*   6-min presentation on Tuesday, June 17 (during lab)

    *   aim for 5 min + 1ish min for Q&A
    
    *   make sure you do not exceed 6 min (practice and time yourself)
    
    *   I'll cut you off if you exceed 6 min

## Capstone project presentation (a preview)

* 2 presentation checkpoints (both during lab time)

* 1 final presentation on the final day (July 25) (+ other deliverables)

* Details will be provided soon

# Presentation tips

## Begin with background and motivation

Every presentation has a **story**:

  - What is the motivation? Why should people care about your work?
  
  - You want to build up what your work is trying to address

. . .

Example: [Ron's nflWAR talk at NESSIS 2017](https://youtu.be/djD-yL3vWNQ):

  - Do **NOT** begin with: "We're introducing [project topic]" (WAR for NFL)
  
  - Instead begin with current state of NFL analytics and need for better, reproducible player level-metrics

. . .
  
Do **NOT** include an outline slide!

  - Your presentation should flow naturally 
  
. . .

[Or you can begin with a "creative" intro...](https://youtu.be/_MAs16nqVdw)


## Describing the data

You want to provide a general overview of your dataset:

  - What are your observations? i.e., what does each row of your dataset represent?

. . .
  
  - What are the relevant variables/features? i.e., what are the columns of interest?
  
    - Be careful though with many variables - avoid just listing everything!
    
    - Simplify by describing groups of variables together
    
. . .

**Use examples** - makes your data explicit and concrete for the audience

  - But do **NOT** print out raw `R` console output or screenshots of output!
  
  - Use text or a clean formatted table (via e.g., `knitr`, [`DT`](https://rstudio.github.io/DT/), [`gt`](https://gt.rstudio.com/), etc.)
  

## Introducing and describing methods

- Prior to presenting results, you want to clearly state any transformations and methods used in the analysis

- Your presentation should provide the general steps for someone to replicate your work

  - e.g., Used complete-linkage hierarchical clustering with [INSERT VARIABLES], determined $k$ number of clusters by [INSERT REASON]
  
  - e.g., Modeled [INSERT RESPONSE VARIABLE] as a function of [INSERT EXPLANATORY VARIABLES]
  
. . .

- For more complicated methods, you'll want to provide a brief review of the methodology

- If introducing new methodology: **walk through the steps clearly**

. . .

- **Always justify your choice of methodology**

  - Why you used a flexible tree-based model over linear regression
  


## Presenting results

- Use the [**assertion-evidence**]( https://www.cmu.edu/student-success/other-resources/handouts/comm-supp-pdfs/designing-powerpoint-slides.pdf) model

- **Assertion**: title of the slide should be the key takeaway in brief sentence form

  - Indicates the point of the visualization or whatever means used to display the results

- **Evidence**: the body of the slide containing the results

  - Display of the results in some format that is simple to explain and understand
  
. . .

- Limit the amount of text in your **Evidence** portion - brief statements with important context

- Treat the **Assertion** as the title of your **Evidence**

  - Plot titles are then redundant and not necessary with an effective assertion



## Assertion

<center>

<br>

<font size="+10">

Evidence

</font>

(e.g., plots, animations, tables, etc.)

<br>

</center>


## MLB strikeout rates have been increasing throughout history

```{r}
#| echo: false
#| fig-align: center

library(tidyverse)
library(Lahman)
Batting |> 
  filter(lgID %in% c("AL", "NL")) |> 
  group_by(yearID) %>%
  summarize(total_sos = sum(SO, na.rm = TRUE),
            total_atbats = sum(AB, na.rm = TRUE)) |> 
  mutate(so_rate = total_sos / total_atbats) |> 
  ggplot(aes(x = yearID, y = so_rate)) +
  geom_line(color = "darkblue") +
  geom_point(alpha = 0.8, color = "darkblue") +
  labs(x = "Year", y = "Strikeouts / At-bats") +
  theme_light() +
  theme(strip.background = element_blank(),
        text = element_text(size = rel(4)))
```

(Explain the elements of your graph - what is each axis, color, shape, etc referring to? And what is the unit scale?)

## Discussion (and ending a presentation)

- Conclude with a recap of the main points of your work

. . .

- Then point out **limitations** and indicate **future directions**

. . .

- Either end with the **Discussion** slide (or **Acknowledgements** but this is sometimes placed at the beginning)

- Never end a presentation with lone *Thank you* slide! 

  - Either want the audience to focus on the final points in your **Discussion** slide
  
  - Or show a plot that stood out from your presentation (pretty picture!)
  
. . .

- Include back-up **Appendix** slides with additional info, ready for questions

. . .

- Slides for **References** should not be displayed during your talk

  - Their purpose is just for sharing with others
  
  - Alternative option: include references directly on slides either in text or via footnotes ^[Like this...]




## Additional tips and reminders

**Use pauses effectively** to highlight points and explain steps

. . .

- Showing all of your text at once can overwhelm your audience

. . .

But

. . .

don't

. . .

be 

. . .

ridiculous

. . .

Remember: **memory overload is real!**

- Do **NOT** introduce too much notation at once

- **Repetitive language and usage of words are useful and reminders for the audience**

  - Use consistent language and terminology throughout the talk

. . .

**Know your audience!**

## Design tips

* Less ink, less ink, less ink

* Your plot should be big enough (font size, line width, point size, etc.)

* Reformat variable names (don't show the original names in the data)

* If a table can be represented by a figure, turn it into a figure (e.g., comparison of model evaluation metrics)

## More general tips

> Your slides should support what you say, not create interference between speech and vision brain areas -- your slides serves as a summary of what you said so someone who may have been distracted can catch up quickly.

> Do not create your talk at the last minute (it will be obvious if you do). Designing good slides takes time. Practicing a talk takes time. 

* [Advices from Andrew Patton:](https://x.com/anpatt7/status/1746209633562443948)

> You don’t need 17 decimal points for anything ever.

> Make sure your figures and charts, etc. all display the correct size. If I have to zoom in 27 times to read the legends that’s not going to help your case.

> Pick better colors for your charts. Your submission should look expensive.

# Do it live: presentation critique

# Do it live: making presentations with Quarto

## Resources to follow along

* Quarto presentation: [`quarto.org/docs/presentations`](https://quarto.org/docs/presentations/)

* `revealjs`: [`quarto.org/docs/presentations/revealjs`](https://quarto.org/docs/presentations/revealjs/)