---
title: "Exploring Chocolate Characteristics and Ratings"
author: "Erin Franke and Sara Colando"
date: Jun 12, 2026
date-format: long
#footer:  "Carnegie Mellon University, Department of Statistics & Data Science"
bibliography: critique-references.bib
format: 
  revealjs:
    theme: theme.scss
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.5
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
    include-after-body: 
      - text: |
          <script type="text/javascript">
          Reveal.addEventListener('slidechanged', (event) => {
            const isSnOn = (event.currentSlide.dataset.hideSlideNumber !== 'true');
            Reveal.configure({ slideNumber: isSnOn ? 'c/t' : false });
          });
          </script>
---

```{r}
#| echo: false
#| fig-height: 9
library(tidyverse)
chocolate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-01-18/chocolate.csv') |>
  mutate(cocoa_percent = as.numeric(str_remove(cocoa_percent, "%")), 
         n_ingredients = as.numeric(str_sub(ingredients, start = 1, end =1)))

list_of_adjectives <- chocolate %>%
  mutate(id = row_number(), # gives each reviewer/chocolate combination a unique id
         most_memorable_characteristics = strsplit(as.character(most_memorable_characteristics), ",")) %>% # split on the comma to create a list of characteristics for each individual
  unnest(most_memorable_characteristics) # unlists the list, one in each column

word_ratings <- list_of_adjectives |>
  group_by(most_memorable_characteristics) |>
  summarize(n = n(), avg_rating = mean(rating))
```

## Overview 

* In this presentation, we are analyzing chocolate ratings around the world. 

* Chocolate is important because it is a favorite food of many people around the world. 

* We use a dataset that contains expert reviews of specialty chocolate bars from manufacturers around the world.

* Each observation represents an individual chocolate bar review identified by a unique reference number.

* We want to explore what makes a chocolate bar highly rated. 

* To do so, we will use plots and clustering. 

* We find rich and creamy chocolates are most highly rated. 

## Chocolate: the world's most popular sweet treat

* The history of chocolate began with the Maya, who were probably the first people in South America to cultivate the cocoa plant [@verna].

. . . 

* Since then, it has become a remarkably versatile ingredient, being used in drinks, desserts, beverages, and even savory food. 

. . . 

<br>

<div style="text-align: center;">**Research Question**: What are the characteristics of highly rated chocolate?</div>

## Getting into the data

2530 rows, 10 variables (relevant ones shown below)

```{r}
#| echo: true
#| fig-height: 9
chocolate |> 
  head() |>
  select(ref, country_of_bean_origin, cocoa_percent, most_memorable_characteristics, rating)
```

## Data: Expert chocolate reviews from 2006-2021

```{r}
#| echo: false
#| fig-height: 9
library(gt)
chocolate |> 
  slice_sample(n=6) |>
  select(ref, review_date, company_manufacturer, country_of_bean_origin, cocoa_percent, most_memorable_characteristics, rating, n_ingredients) |>
  gt() |>
  cols_label(ref = "Review ID",
    review_date = "Year", 
    company_manufacturer = "Manufacturer",
    country_of_bean_origin = "Country bean of origin", 
    cocoa_percent = "Cocoa %", 
    most_memorable_characteristics = "Adjectives", 
    rating = "Rating (out of 5)", 
    n_ingredients = "Number of Ingredients")
```

## Scatterplot {data-hide-slide-number="true"}

No relationship between cocoa percentage and rating. 

```{r}
#| echo: false
#| fig-height: 9
chocolate |>
  ggplot(aes(x=cocoa_percent, y=rating))+
  geom_point(alpha = 0.5) 
```

## Cocoa percentage is not indicative of rating 

```{r}
#| echo: false
#| fig-height: 7
#| fig-align: center
chocolate |>
  ggplot(aes(x=cocoa_percent/100, y=rating))+
  geom_jitter(alpha = 0.3, size = 1.5, color = "#4f412b")+
  geom_smooth(color = "skyblue3", fill = "skyblue3", alpha = 0.2)+
  theme_bw()+
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Cocoa Percentage",  y= "Rating")+
  theme(axis.title = element_text(color = "black", size = 18), 
        axis.text = element_text(color = "black", size = 16))
```

## Adjectives describing chocolate versus deviation from the mean rating {data-hide-slide-number="true"} 

```{r}
word_ratings |>
  filter(n>=50) |>
  mutate(overall_avg_rating = mean(chocolate$rating),
         rating_diff = avg_rating - overall_avg_rating) |>
  ggplot(aes(x=rating_diff,y = most_memorable_characteristics))+
  geom_col()
```

## Chocolates described as creamy and rich receives the highest ratings, on average

```{r}
overall_avg_rating <- mean(chocolate$rating)

word_ratings |>
  filter(n>=50) |>
  mutate(overall_avg_rating = mean(chocolate$rating),
         rating_diff = avg_rating - overall_avg_rating) |>
  arrange(desc(rating_diff)) |>
  slice(c(1:5, (n()-4):n())) |>
  mutate(most_memorable_characteristics = fct_reorder(most_memorable_characteristics, rating_diff),
         diff_pos = if_else(rating_diff >= 0, TRUE, FALSE)) |>
  ggplot(aes(x=rating_diff,y = most_memorable_characteristics))+
  geom_col(aes(fill = diff_pos)) +
  geom_vline(xintercept = 0) +
  scale_fill_manual(values=NatParksPalettes::natparks.pals("DeathValley")) +
  theme_bw() +
  labs(x = "Deviation from mean chocolate rating",  y= "Most memorable characteristics")+
  theme(legend.position = "none",
        axis.title = element_text(color = "black", size = 18), 
        axis.text = element_text(color = "black", size = 16))
```

## Clustering chocolate {data-hide-slide-number="true"}

* We ran a clustering algorithm with $k=3$. We find cluster 1 has the highest mean rating at 3.23, whereas cluster 2 has a mean rating of 3.06 and cluster 3 a mean rating of 2.96. 

:::{columns}

:::{.column}

```{r}
#| echo: false
#| fig-height: 9
choc_clusters <- chocolate |> 
  dplyr::select(cocoa_percent, n_ingredients) |>
  mutate(cocoa_percent_sd = as.numeric(scale(cocoa_percent)), 
         n_ingredients_sd = as.numeric(scale(n_ingredients))) |>
  drop_na() |>
  #slice_sample(n=100) |>
  dist() |>
  hclust(method = "complete")

chocolate |> 
  filter(!is.na(cocoa_percent), !is.na(n_ingredients)) |> 
  mutate(
    cluster = as.factor(cutree(choc_clusters, k = 3))
  ) |>
  ggplot(aes(x = cocoa_percent, y = n_ingredients,
             color = cluster)) +
  geom_point(size = 4) + 
  ggthemes::scale_color_colorblind() +
  theme_bw() + 
  theme(legend.position = "bottom")
```

:::

:::{.column}
```{r}
require("ggdendro")

ggdendrogram(
  choc_clusters,
  rotate = FALSE,
  theme_dendro = TRUE
) +
  ggtitle("Chocolate Dendrogram")
```

:::

:::

## Identifying varieties of chocolate with clustering

We performed **hierarchical clustering** with: 

  1. The number of ingredients
    
  2. Cocoa percentage  
  
. . .  
  
Among the **three** clusters, chocolate in *Cluster 1 has the highest average rating*  

```{r}
#| echo: false
#| fig-height: 9
chocolate |> 
  filter(!is.na(cocoa_percent), !is.na(n_ingredients)) |> 
  mutate(
    Cluster = as.factor(cutree(choc_clusters, k = 3))
  ) |>
  group_by(Cluster) |>
  summarize(`Average Cocoa Percent (%)` = mean(cocoa_percent),
            `Average Number of Ingredients` = mean(n_ingredients),
            `Average Rating` = mean(rating)) |>
  gt() |>
  fmt_number(decimals = 2)
```

## Discussion 

* Despite chocolate often being characterized by its cocoa percentage, we do not find cocoa percentage to be indicative of rating. 

. . . 

* Rather, chocolates described with memorable characteristics of *rich* and/or *creamy*, among others, were rated highest.

. . . 

* Future work may explore what aspects of these chocolates make them so delectable.

# [Thanks!]{style="color: red; background-color: #00F5FF; font-family: 'Comic Sans MS', cursive;"} {background-color="#00F5FF"}

