---
title: "EDA project: Women's World Cup passing"
---

---

## Overview

This project will be released on Thursday, June 4 and **conclude with a 6-minute presentation on Tuesday, June 16 during lab**.

Students will be randomly placed into groups of 2--3 and each group will be randomly assigned a dataset.

The goal of this project is to practice understanding the structure of a dataset, and to practice generating and evaluating hypotheses using fundamental EDA and data visualization techniques.

## Deliverables

Each group is expected to make slides to accompany the 6-minute presentation. All group members must be present for the presentation and participate. **No notes/scripts are allowed during the presentation.**

The presentation should feature the following:

* Overview of the structure of your dataset

* **Two questions/hypotheses** you are interested in exploring

* **Two data visualizations** exploring the questions---both must be multivariate (i.e., involving 2+ variables) and in different formats

* **One clustering analysis**

* Conclusions for the hypotheses based on your EDA and data visualizations

## Timeline

There will be two submission deadlines:

**Thursday, June 11 at 5pm ET** - Each student will push their individual code for the project thus far to GitHub for review.

**Tuesday, June 16 at 1pm ET** - Slides and full code must be completed and ready for presentation. Send your slides to Erin ([`efranke@andrew.cmu.edu`](mailto:efranke@andrew.cmu.edu)) and Sara ([`scolando@andrew.cmu.edu`](mailto:scolando@andrew.cmu.edu)). All code must be written in `R`; but the slides may be created in any software. Take advantage of examples from lectures, but also feel free to explore online resources that may be relevant. (But be sure to always consult the `R` help documentation first before attempting to google around or ask ChatGPT.)

## Data

This dataset contains all (completed) pass attempts from the 2023 FIFA Women's World Cup, courtesy of [StatsBomb](https://statsbomb.com/news/statsbomb-release-free-2023-womens-world-cup-data/) and accessed via the [`StatsBombR`](https://github.com/statsbomb/StatsBombR) package.

Each row in the dataset corresponds to a pass attempt and the columns are:

* `period`: period of the game (`1`: first half, `2`: second half, `3`: first extra time period, `4`: second extra time period)          
* `minute` & `second`: minute & second of the game
* `duration`: duration of the pass (seconds)            
* `under_pressure`: whether the pass is under pressure    
* `counterpress`: whether the pass is under a counterpress event (defined as a pressing action occurring within five seconds of an open play turnover)  
* `play_pattern.name`: play pattern
* `team.name`: team name
* `player.name`: passer name  
* `position.name`: passer position
* `pass.{switch, aerial_won, cross, through_ball, ..., no_touch, deflected, miscommunication}`: different pass binary indicators (should be self-explanatory, `NA` means `FALSE`)
* `pass.recipient.name`: pass recipient name
* `pass.height.name`: type of pass height
* `pass.body_part.name`: body part used to make the pass
* `pass.type.name`: type of pass
* `pass.outcome.name`: outcome of the pass 
* `pass.technique.name`: technique of the pass
* `location.{x, y}`: (`x`, `y`) location when pass takes place. (Note: all locations have been standardized so that the possession team is going from left to right. Higher `x` values means closer to the opposing team's goal, and higher `y` values means more toward the left wing)
* `pass.end_location.{x, y}`:  (`x`, `y`) coordinates of pass ending location
* `pass.length`: length of the pass
* `pass.angle`: angle of the pass
* `TimeInPoss`: time elapsed since the start of possession         
* `TimeToPossEnd`: time remaining until end of possession

## Starter code

```{r}
#| eval: false
library(tidyverse)

wwc_passes <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/wwc_passes.csv")
```

In case you're curious, the code to build this dataset can be found below.

```{r}
#| eval: false
library(tidyverse)
library(StatsBombR)

wwc <- FreeCompetitions() |> 
  filter(competition_id == "72" & season_id == "107") |> 
  FreeMatches() |> 
  free_allevents() |> 
  allclean()

wwc_passes <- wwc |>
  filter(type.name == "Pass", !is.na(pass.recipient.name)) |> 
  janitor::remove_empty() |> 
  select(period, minute, second, duration, under_pressure, counterpress, 
         play_pattern.name, team.name, player.name, position.name, 
         pass.switch:pass.through_ball, pass.shot_assist:pass.goal_assist, 
         pass.cut_back:pass.miscommunication, pass.recipient.name, pass.height.name, 
         pass.body_part.name, pass.type.name, pass.outcome.name, pass.technique.name,  
         location.x, location.y, pass.end_location.x, pass.end_location.y, 
         pass.length, pass.angle, TimeInPoss, TimeToPossEnd)
```